Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2016-03-15 rmarkdown/paper.Rnw
fixed background section of paper

ChIP-exo (Chromatin Immunoprecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
\cite{exo1} is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments that the protein under study interacts with using random
fragmentation of DNA and a protein-specific antibody. Then,
exonuclease is introduced to trim the $5\prime$ end of each DNA
fragment to a fixed distance from the bound protein compared to
ChIP-Seq. This step is unique to ChIP-exo and could potentially
provide significantly higher spatial resolution compared to
ChIP-Seq. Finally, high throughput sequencing of a small region (36 to
100 bp) at the $5\prime$ end of each fragment generates millions of
reads. Figure \ref{fig:chip_diagram} illustrates the differences
between ChIP-exo, Single End (SE) ChIP-Seq and Paired End (PE)
ChIP-Seq: The $5\prime$ ends of a ChIP-exo experiment are located more
tightly around the binding proteins than in a ChIP-Seq experiment; in
a PE ChIP-Seq experiment both ends are observed while in a SE ChIP-Seq
experiment only the $5\prime$ end.

@@

While the number of ChIP-exo data keeps increasing, characteristics of
ChIP-exo data are not fully investigated yet. First, DNA libraries
generated by the ChIP-exo protocol seem to be less complex than the
libraries generated by ChIP-Seq (Mahony et al., 2015,
\cite{exo_review}), i.e. the possible number of positions to which the
reads can be aligned has been reduced due to the exonuclease
digestion. Second, although there are roughly the same amount of read
in both strands, locally there may be more reads in one strand than in
the other.

To address these challenges, we suggest a collection of
quality control visualizations to interrogate these biases in a
ChIP-exo experiment and globally asses the enrichment and library
complexity of a ChIP-exo sample. We gathered ChIP-exo data from
diverse organisms: CTCF factor in human \cite{exo1}; ER factor in
human and FoxA1 factor in mouse (Serandour et al., 2013
\cite{exoillumina}); Glucocorticoid receptor (GR) in IMR90, K562 and
U20S cell lines (Starick et al., 2015 \cite{starick15}).Furthermore,
we also generated ChIP-exo and ChIP-Seq data for $\sig$ factor in
\emph{Escherichia Coli} (\emph{E. Coli}) under aerobic ($ + O_2$)
condition, and treated by rifampicin by 0 and 20 minutes (courtesy of
Professor Robert Landick's lab).
%% note to self, we may add more experiments here

Most of current ChIP-Seq quality control (QC) guidelines (Landt et
al., \cite{encode_qc}) may not be applicable on ChIP-exo, additionally
to our knowledge there are not established QC pipelines for
ChIP-exo. Previous ChIP-exo analyses used ChIP-Seq samples to compare
the resolution between experiments; Carroll et al., 2014
\cite{carroll.qc} studied the use of the Strand-Cross Correlation
(SCC) (Kharchenko et al., 2008 \cite{strandcc}) and showed that by
filtering blacklisted regions the estimation of the SCC is
improved. However, using the SCC may not be helpful since the peaks
that are attained at the read and fragment lengths are confused in a
typical ChIP-exo SCC curve. Additionally this method requires to know
blacklisted regions in advance which may not be available. In our
pipeline we propose an out-the-shelf analysis that contrasts the
enrichment of the experiment against its library complexity.

In order to obtain the potential benefits of ChIP-exo on protein
binding site identification, it is critical to understand which are
the important characteristics of ChIP-exo data and to use algorithms
that could fully utilize information available in ChIP-exo data. Rhee
and Pugh, 2011 \cite{exo1} discussed that reads in the forward and
reverse strand might construct peak pairs around bound proteins, of
which heights were implicitly assumed to be symmetric. Hence, they
used the ``peak pair method'' that predicts the midpoint of two modes
of peak pairs as potential binding sites. Mace (Wang et al, 20114
\cite{mace}), CexoR (Madrigal, 2015 \cite{cexor}) and Peakzilla
(Bardet et al., 2013 \cite{peakzilla}), recently developed ChIP-exo
data analysis methods are also based on this peak pair
assumption. However, appropriateness of such assumption was not fully
evaluated in the literature yet. Furthermore, it is still unknown
which factors could affect protein binding site identification using
ChIP-exo data. In order to address this problem, we investigated
various aspects of ChIP-exo data by contrasting them with their
respective ChIP-Seq experiments.

Currently, research on statistical methods for ChIP-exo data is still
in its very early stage. Although many methods have been proposed to
identify protein binding sites from ChIP-Seq data (reviewed by
Wilbanks and Facciotti, 2012 \cite{evaluation} and Pepke and Wold,
2009 \cite{computation}), such as MACS (Zhang et al., 2008
\cite{macs}), CisGenome (Ji et al., 2008 \cite{cisgenome}) and MOSAiCS
(Kuan et al., 2009 \cite{mosaics}), these approaches reveal protein
binding sites in lower resolution, i.e., at an interval of hundreds to
thousands of base pairs. Furthermore, they report only one ``mode'' or
``predicted binding location'' per peak. Hence, these methods are not
appropriate to evaluate the potential of ChIP-exo data for high
resolution identification of protein binding sites. More recently,
deconvolution algorithms such as Deconvolution (Lun et al., 2009
\cite{csdeconv}), GEM (Guo et al., 2012 \cite{gem}, an improved
version of Guo et al., 2010 \cite{gps} ) and PICS (Zhang et al., 2010
\cite{pics}) have been proposed to identify binding sites in higher
resolution using ChIP-Seq data. However, most of them are still not
tailored for ChIP-exo and PE and SE ChIP-Seq data in a unified
framework and as a result, currently available methods are not
appropriate for fair comparison between ChIP-exo and ChIP-Seq. To
address these limitations, we developed an improved of dPeak (Chung et
al., 2013 \cite{dpeak}), a high resolution binding site identification
(deconvolution) algorithm that we previously developed for PE and SE
ChIP-Seq data, so that it can also handle ChIP-exo data. The dPeak
algorithm implements a probabilistic model that accurately describes
the ChIP-exo and ChIP-Seq data generation process.

In this work we demonstrate that the ``peak pair'' assumption of Rhee
and Pugh, 2013 \cite{exo2} does not hold well in real ChIP-exo
data. Furthermore, we found that when we analyze ChIP-exo data from
eukaryotic genomes, it is important to consider sequence biases
inherent to ChIP-exo data, such as mappability and GC content in order
to improve sensitivity and specificity of binding site
identification. We evaluated several methods to identify binding events
and dPeak performs competitively respect to GEM and MACE when
analyzing ChIP-exo data. More importantly, when comparable number of
reads is used for both ChIP-exo and ChIP-Seq , dPeak coupled with
ChIP-exo data provides resolution comparable to PE ChIP-Seq and both
significantly improve the resolution of protein binding identification
compared to SE-based analysis with any of the available methods.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width =
  .95 \textwidth]{/p/keles/ChIPexo/volume3/ChIPexo/figs/for_paper/chip_explanation_withEnzyme.pdf}
  \caption{\textbf{Description of ChIP-exo, SE ChIP-Seq and PE
      ChIP-Seq:} The DNA is sonicated, for ChIP-exo a exonuclease
    enzyme (orange hexagon) trims the $5\prime$ ends of each DNA
    fragment to a fixed distance from the bound protein, finally is
    subjected to Immunoprecipitation and amplification. For both
    ChIP-exo and SE ChIP-Seq an adapter is ligated (green triangles)
    at the $5\prime$ ends, while for PE ChIP-Seq is ligated to both
    ends.}
  \label{fig:chip_diagram}
\end{figure}
\newpage

>>>>

ChIP-exo (Chromatin Immunoprecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
\cite{exo1} is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments (150 - 250 bp in length) that the protein under study
interacts with using random fragmentation of DNA and a
protein-specific antibody. Then, exonuclease is introduced to trim the
$5\prime$ end of each DNA fragment to a fixed distance from the bound
protein compared to ChIP-Seq. This step is unique to ChIP-exo and
could potentially provide significantly higher spatial resolution
compared to ChIP-Seq. Finally, high throughput sequencing of a small
region (36 to 100 bp) at the $5\prime$ end of each fragment generates
millions of reads. Figure \ref{fig:chip_diagram} illustrates the
differences between ChIP-exo, Single End (SE) ChIP-Seq and Paired End
(PE) ChIP-Seq: The $5\prime$ ends of a ChIP-exo experiment are located
more tightly around the binding proteins than in a ChIP-Seq
experiment; in a PE ChIP-Seq experiment both ends are observed while
in a SE ChIP-Seq experiment only the $5\prime$ end.

@@

While the number of ChIP-exo data keeps increasing, characteristics of
ChIP-exo data are not fully investigated yet. First, DNA libraries
generated by the ChIP-exo protocol seem to be less complex than the
libraries generated by ChIP-Seq (Mahony et al., 2015
\cite{exo_review}). Second, although there are roughly the same amount
of reads in both strands, locally there may be more reads in one
strand than in the other. Finally, most of current ChIP-Seq quality
control (QC) guidelines (Landt et al., \cite{encode_qc}) may not be
applicable on ChIP-exo, while there are not established QC pipelines
for ChIP-exo; previous ChIP-exo analyses used ChIP-Seq samples to
compare the resolution between experiments (\cite{exo1},
\cite{exoillumina}, \cite{exo2})). To address these challenges, we
suggest a collection of quality control visualizations to interrogate
these biases in a ChIP-exo experiment and globally asses the
enrichment and library complexity of a ChIP-exo sample. We gathered
ChIP-exo data from diverse organisms: CTCF factor in human
\cite{exo1}; ER factor in human and FoxA1 factor in mouse (Serandour
et al., 2013 \cite{exoillumina}); Glucocorticoid receptor (GR) in
IMR90, K562 and U20S cell lines (Starick et al., 2015
\cite{starick15}).Furthermore, we also generated ChIP-exo and ChIP-Seq
data for $\sig$ factor in \emph{Escherichia Coli} (\emph{E. Coli})
measured under aerobic ($ + O_2$) condition, and treated by rifampicin
by 0 and 20 minutes (courtesy of Professor Robert Landick's lab).

In order to obtain the potential benefits of ChIP-exo on protein
binding site identification, it is critical to use algorithms that
could fully utilize information available in ChIP-exo data. Rhee and
Pugh, 2011 \cite{exo1} discussed that reads in the forward and reverse
strand might construct peak pairs around bound proteins, of which
heights were implicitly assumed to be symmetric. Based on this
rationale, they used the ``peak pair method'' that predicts the
midpoint of two modes of peak pairs as potential binding
sites. Recently developed ChIP-exo data analysis methods, such as Mace
(Wang et al, 20114 \cite{mace}), CexoR (Madrigal, 2015 \cite{cexor})
and Peakzilla (Bardet et al., 2013 \cite{peakzilla}), are also based
on this peak pair assumption. However, appropriateness of such
assumption was not fully evaluated in the literature yet. Furthermore,
it is still unknown which factors could affect protein binding site
identification using ChIP-exo data. In order to address this problem,
we investigated various aspects of ChIP-exo data by contrasting them
with their respective ChIP-Seq experiments.

Currently, research on statistical methods for ChIP-exo data is still
in its very early stage. Although many methods have been proposed to
identify protein binding sites from ChIP-Seq data (reviewed by
Wilbanks and Facciotti, 2012 \cite{evaluation} and Pepke and Wold,
2009 \cite{computation}), such as MACS (Zhang et al., 2008
\cite{macs}), CisGenome (Ji et al., 2008 \cite{cisgenome}) and MOSAiCS
(Kuan et al., 2009 \cite{mosaics}), these approaches might not fully
utilize potentials of ChIP-exo data for high resolution identification
of protein binding sites. Specifically these approaches reveal protein
binding sites only in lower resolution, i.e., at an interval of
hundreds to thousands of base pairs. Furthermore, they implicitly
assume that there is only one ``mode'' or ``predicted binding
location'' per this wide genomic interval. More recently,
deconvolution algorithms such as Deconvolution (Lun et al., 2009
\cite{csdeconv}), GEM (Guo et al., 2012 \cite{gem}, an improved
version of Guo et al., 2010 \cite{gps} ) and PICS (Zhang et al., 2010
\cite{pics}) have been proposed to identify binding sites in higher
resolution using ChIP-Seq data. However, most of them are still not
tailored for ChIP-exo and PE and SE ChIP-Seq data in a unified
framework and as a result, currently available methods are not
appropriate for fair comparison between ChIP-exo and ChIP-Seq. To
address these limitations, we developed and utilized an improved
version of dPeak (Chung et al., 2013 \cite{dpeak}), a high resolution
binding site identification (deconvolution) algorithm that we
previously developed for PE and SE ChIP-Seq data, so that it can also
handle ChIP-exo data. The dPeak algorithm implements a probabilistic
model that accurately describes the ChIP-exo and ChIP-Seq data
generation process.

Some of the key findings in this work are as follows. First, we
demonstrate that the ``peak pair'' assumption of Rhee and Pugh, 2013
\cite{exo2} does not hold well in real ChIP-exo data. Second, we found
that when we analyze ChIP-exo data from eukaryotic genomes, it is
important to consider sequence biases inherent to ChIP-exo data, such
as mappability and GC content, in order to improve sensitivity and
specificity of binding site identification. Third, we evaluated
several methods to identify binding events and dPeak performs
competitively respect to GEM and MACE when analyzing ChIP-exo
data. Finally, when comparable number of reads is used for both
ChIP-exo and ChIP-Seq , dPeak coupled with ChIP-exo data provides
resolution comparable to PE ChIP-Seq and both significantly improve
the resolution of protein binding identification compared to SE-based
analysis with any of the available methods.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width =
  .95 \textwidth]{/p/keles/ChIPexo/volume3/ChIPexo/figs/for_paper/chip_explanation_withEnzyme.pdf}
  \caption{\textbf{Description of ChIP-exo, SE ChIP-Seq and PE
      ChIP-Seq:} A TF is bound to the forward (red) and backward
    (blue) strand of the DNA. Then, is sonicated: For ChIP-exo a
    exonuclease enzyme (orange hexagon) trims the $5\prime$ ends of
    each DNA fragment to a fixed distance from the bound protein,
    finally is subjected to Immunoprecipitation and amplification. For
    both ChIP-exo and SE ChIP-Seq an adapter is ligated (green
    triangles) at the $5\prime$ ends, while for PE ChIP-Seq is ligated
    to both ends.}
  \label{fig:chip_diagram}
\end{figure}
\newpage
